---
title: 'HW2-EDA with R - Part1: Plotting'
author: "misken"
date: "January 22, 2018"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Familiarize yourself with the data and the assignment

In this assignment you'll do some exploratory data analysis
with R on a dataset US census information based on the [American Community Survey](https://www.census.gov/programs-surveys/acs/). The version of
this data that we'll be using is from a Kaggle dataset. See
https://www.kaggle.com/muonneutrino/us-census-demographic-data/data and look
at the column metadata for the county level table to see what the various
columns mean.

As we did in class, you'll be creating an R Markdown document to
both prepare a dataset and do some analysis as well as to document the
steps you did (and answer some questions I'll throw at you).

You'll notice a few "Hacker Extra" tasks
thrown in. These are for those of you who want to go a little above and beyond
and attempt some more challenging tasks. And, feel free to do a some free form
Hacker Extra style work yourself - in other words, do something beyond what
was asked. You'll learn more.

## Step 2 - Create a new R Markdown document

Save this file as a new R Markdown document and name it **HW2_EDAwR_[_your last name_].Rmd**. Mine would
be called **HW2_EDAwR_isken.Rmd**. Save it into the same folder as this file. 

This will be your working directory. 

Create a new R Project based on this working directory.

**Hacker Extra** - Put this project folder under version control using git. Only
track the .R file, the .Rmd file and your .gitignore file. Create a remote repo
for this homework assignment in your GitHub account. Periodically push your commits
to GitHub. When done with the assignment, paste in the URL of your GitHub project
so that I can go look at it.

## Step 3 - Set working directory and explore data folder

Set the working directory or create an R Studio project in the current folder. You'll notice that there is a folder named **data**.
Inside of it you'll find the data files for this assignment:

- **acs2015_census_tract_data.csv**
- **acs2015_county_data.csv**
- **us_census_bureau_regions_and_divisions.csv**
- **county.csv**
- **county_MI.csv**

The first two files are from the Kaggle site for this dataset. The two files,
**county.csv** and **county_MI.csv**, are files that I created. The first
contains county level data from all states and the other just data from
Michigan. The R script, **census_data_prep.R** was used to create these files
and to merge information from the **us_census_bureau_regions_and_divisions.csv**
file with the county data.

You might want to use shell commands or a text editor like Geany to snoop the files a bit.


## Step 4 - Complete the following R tasks and answer questions

Now you'll need to complete the following tasks in R. Just like we did in class, you should
use a combination of markdown text (be concise, no need to write tons of text) 
to explain what you are doing and R code chunks to actually do it.
When you are done, use the "Knit" button to generate an HTML file from your R Markdown.
You'll be submitting BOTH the completed R Markdown file as well as the generated HTML file. You'll notice that I've left some empty code chunks
as well as some partially filled in code chunks. Of course, feel free to
add more code chunks as needed.

Let's load a few libraries we'll need:

```{r}
library(dplyr)
library(ggplot2)
```


### Problem 1 - Get the data into a dataframe

You should read the **county.csv** file into a data frame named county_df and
**county_MI.csv** into a data frame named county_MI.

```{r read_countydata}
county_df <- read.csv('data/county.csv')
county_MI <- read.csv('data/county_MI.csv')

```

Check the structure of each data.frame and use the `summary` function to begin
exploring the data. Are there any NA values? What 
differences do you notice between the structure of the two data frames.

*ANSWER: There are NA values in the following variables for county_df: "Income" "IncomeErr" "ChildPoverty". There are no NA values in county_df. The dataframe county_MI has 41 variables, while county_df only has 40 variables. Because county_MI only contains observations from Michigan, the variable "peninsula" was added to specify if the observation was from the lower peninsula or the upper peninsula. Another difference between the two structures is that county_MI has less levels in its factor variables because the observations in this data frame are only from Michigan.*

```{r}
str(county_df)
summary(county_df)
str(county_MI)
summary(county_MI)
```

The county_MI data frame contains data for Michigan only and includes an
peninsula field. See **census_data_prep.R** to see how this field was
created. It involves scraping a table out of a web page.

Unless otherwise specified, you should use the ggplot2 package for all of the
following plots.

### Problem 2 - Basic plots for Income

Create a histogram, boxplot, kernel density plot and violin plot for median
income for the entire US. 

Write a short paragraph describing the distribution of median incomes.

*The distribution has a wide spread with median income ranging from less than $25,000 in some counties to almost $125,000 in others.The distribution of median income is unimodal and skewed to the right with a mode of about $40,000. The distribution also varies based on the region or division. Median income distribution in the New England and Pacific regions have a much more narrow distribution and have significantly fewer outliers than the other divisions. Similarly, the Midwest region is much more narrowly distributed than the South region*
**

```{r}
ggplot(county_df, aes(county_df$Income)) + 
  geom_histogram(col = "black", fill = "blue") + 
  labs(title = "US Median County Income", x = "Median Income", y = "Number of Counties in US")
ggplot(county_df, aes(x = county_df$Division, y = county_df$Income)) + 
  geom_boxplot(col = "black", fill = "green") + 
  labs(title = "US Median County Income by Division", x = "Division", y = "Median County Income") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
ggplot(county_df, aes(county_df$Income, fill = county_df$Region)) + 
  geom_density(position = "stack", alpha = .4) + 
  labs(title = "US Median County Income by Region", x = "Median Income", y = "Density", fill = "Region") + 
  scale_y_continuous(labels = scales::comma) + 
  scale_color_brewer(palette = "Accent")
ggplot(county_df, aes(x = county_df$Region, y = county_df$Income)) + 
  geom_violin(col = "pink", fill = "brown") + 
  labs(title = "US Median County Income by Region", x = "Region", y = "Median County Income") + 
  theme(axis.text.x = element_text(angle = 75, hjust = 1))
```

### Problem 3 - Histogram of total population

Create a histogram of total population for counties in the US.

```{r histo_totpop}

```

Repeat the histogram about but make the following style changes to the plot.

* Add a title such as "Distribution of population for US counties"
* Change y-axis label to "Number of counties"
* Change x-axis label to "Total Population"
* Change the colors of the bars to something other than the default black

```{r histo_stylechanges}

```

**Hacker Extra** - Also create a kernel density plot overlaid on a histogram of
Income. 

Repeat the histogram but only for counties with less than 2.5 million people.

```{r histo_lt2.5m}

```


### Problem 4 - How do Michigan's peninsulas compare?

Create a grouped box plot showing Income by peninsula. Make sure
your graph has a title and good axis labels.

Do the peninsulas 
differ with respect to distribution of median income over the counties?


```{r bp_inc_by_peninsula}

```

Create several more boxplots showing how the peninsulas compare with respect to
some of the other variables. Summarize your findings in a few short paragraphs.

```{r more_boxplots}

```

### Problem 5 - How do regions of the country compare?

In county_df you'll see that there's a Region field. Create a set of faceted 
histograms showing how some variable of interest to you compares by region. For 
example, you could explore how the regions compare with respect to the
distribution of MeanCommute time. Make sure your graph has a title and good axis
labels.

```{r faceted_histo}

```

### Problem 6 - Basic bar chart

Create a bar chart showing the number of counties by Region in the US. Make sure
your graph has a title and good axis labels.

```{r bar_counties_by_region}

```

### Problem 7 - Basic scatter plots

Create a scatter plot using the Michigan only data frame comparing
Income and MeanCommute.

```{r xy_inc_commute}

```

Create a modified version of the above scatter plot in which the point color
is mapped to peninsula.

```{r xy_inc_commute_peninsula}

```

Your thoughts on the relationship between these variables?

Now, create a few more scatter plots using variables of your choosing in which
you facet by peninsula as well. How do the relationships compare across the
peninsulas?

```{r xy_faceted_peninsula}

```


Now use the full US data frame.

Create a scatter plot of TotalPop by Income.

```{r xy_pop_inc}

```

The widely varying values make this difficult to interpret in terms of looking
for patterns. So, figure out a way to either change the axis scales or 
the plotted data itself to use a base 10 logarithmic scale (or transform).
Write a short paragraph on why one might want to use a log scale or log
transform of data when plotting.

### Hacker Extra - More analysis

Do some additional plotting to look for interesting patterns or relationships
either in the full US dataset or the MI only dataset. In addition, try out
some of ggplots additional features such as themes, axis formatting, gridline
control, or others.